---
title: "Extending the ensemble stack"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending the ensemble stack}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dev = "ragg_png")
```

```{r setup}
library(ens)
```

`ens` is the foundation of a family of paleogeoscience packages
(`lipdViz` for visualization, `geoChronR` for age modeling, `actR` for
abrupt-change detection, `compositeR` for record compositing). This
vignette describes two shared conventions that keep the family modular
and make it easy to add new methods that reuse existing infrastructure.

## 1. The ensemble-propagation engine

Most time-uncertain analyses follow the same skeleton: take a record,
generate an ensemble that represents its age and/or value uncertainty,
run an analysis on each ensemble member, and (optionally) test the
result against surrogates. `ens` provides that skeleton so you only have
to write the analysis itself.

### The `changeFun` contract

Write a function that takes a single `time` vector and `vals` vector
(plus any parameters) and returns **a one-row data frame**:

```{r}
# detect whether the second half of a record differs from the first
meanShiftFrac <- function(time, vals, split = median(time)) {
  pre  <- vals[time <  split]
  post <- vals[time >= split]
  tibble::tibble(delta = mean(post, na.rm = TRUE) - mean(pre, na.rm = TRUE))
}
```

Any function with this shape automatically gains:

- **uncertainty propagation** via `propagateUncertainty()`, which maps it
  across an ensemble (simulating age uncertainty with `simulateBam()` and
  value uncertainty with `simulateAutoCorrelatedUncertainty()` when an
  ensemble is not supplied), and
- **null-hypothesis testing** via `testNullHypothesis()`, which runs the
  same function on surrogate series.

```{r}
set.seed(1)
time <- seq(0, 1000, by = 10)
vals <- rnorm(length(time)) + ifelse(time > 500, 1.5, 0)  # a shift at 500

# propagate uncertainty: one row per ensemble member
ens_out <- propagateUncertainty(time, vals,
                                changeFun = meanShiftFrac,
                                n.ens = 50,
                                progress = FALSE)
quantile(ens_out$delta, c(0.025, 0.5, 0.975))
```

### Summaries are caller-supplied

`propagateUncertainty()` returns every ensemble member by default. To
collapse the ensemble, pass a *function* describing how to summarize your
own output -- the engine stays agnostic to what your `changeFun` returns:

```{r}
propagateUncertainty(time, vals, changeFun = meanShiftFrac, n.ens = 50,
                     progress = FALSE,
                     summarize = function(p) {
                       tibble::tibble(median_delta = median(p$delta),
                                      frac_positive = mean(p$delta > 0))
                     })
```

### Significance from surrogates

`testNullHypothesis()` runs the analysis on surrogates that preserve the
record's statistical properties but destroy its signal, and `kdePval()`
turns the resulting null distribution into a p-value:

```{r}
nulls <- testNullHypothesis(time, vals, changeFun = meanShiftFrac,
                            n.ens = 20, mc.ens = 20,
                            surrogate.method = "isospectral",
                            progress = FALSE)
null_deltas <- vapply(nulls, function(x) median(x$delta), numeric(1))
observed <- median(ens_out$delta)
kdePval(null_deltas, observed)$pval
```

That is the entire pattern `actR` uses for excursion and shift
detection: each detector is just a `changeFun`.

## 2. The two-layer result convention

Across the family, results follow a deliberate two-layer split:

- **Inside**, heavy numerics stay plain matrices and vectors -- fast,
  memory-light, and easy to compute on.
- **Outside**, user-facing results carry an S3 class with
  `print`/`plot`/`summary` methods, and use a small set of standardized
  component names so functions compose across packages.

The container follows the shape of the data:

| Result shape | Container | Example |
|---|---|---|
| One row per record/ensemble member (row-composable) | classed **tibble** with list-columns | `actR`'s `excursion`/`shift` |
| Matrix-shaped (bins x ensemble) | classed **list** of matrices | `compositeR`'s `paleoComposite` |

Shared component names — `time`, `values`, `ensemble`, `parameters`,
`metadata` — let a result from one package flow into a plotting or
analysis function in another without translation. When you add a new
method, return matrices internally, attach a class for display, and reuse
these names so the rest of the stack already knows how to handle it.
```
